\documentclass{acm_proc_article-sp}

\begin{document}

\title{Extended Related Works of "Fast Counting of Triangles in Large Real Networks without counting:}
\subtitle{Algorithms and Laws"}

\numberofauthors{1}
\author{
\alignauthor
Amanda Wulf\\
       \affaddr{University of Texas at San Antonio}\\
       \affaddr{1 UTSA Circle}\\
       \affaddr{San Antonio, Texas 78240}\\
       \email{amandakwulf@gmail.com}
}

\maketitle
\begin{abstract}
This paper serves as an extended background explanation, motivation, and related
works section for \cite{original}.
\end{abstract}

\section{Introduction}
This paper's goal is to provide an extended background and related works for
\cite{original}. \cite{original} is a paper detailing a method of triangle
estimation in real-world graphs using a mathematical relationship between the
number of eigenvalues of real-world graphs and their triangle count. This paper
introduces the basic ideas of graph theory, the motivations behind this work,
the previous algorithms developed for triangle counting, and insights into why
this particular triangle counting algorithm works.

\section{Graph Theory Background}
A graph G is defined as a way of encoding pairwise relationships among a set of
objects. It consists of a collection of nodes V and a collection of edges E, each
of which joins two nodes. Each edge v is an ordered pair (u, v). An edge e is
said to leave node u and enter node v. Nodes are also called vertices
\cite{kleinberg}.

Graphs can be either directed (meaning an edge can only be traversed in a
certain direction) or undirected \cite{kleinberg}. In this paper, the graphs we
consider are undirected.

A self-edge (or loop) is a node that has an edge connecting it to itself
\cite{diestel}.

Two nodes are adjacent if they have an edge connecting them. A fully
connected graph (also known as a complete graph) is a graph in which all
nodes are pairwise adjacent \cite{diestel}.

A triangle is defined to be a set of three fully connected nodes in an
undirected graph without self-edges \cite{original}.

The degree of a node is defined to be the number of edges that connect to it
\cite{diestel}.

A closed walk of a graph is a finite, loop-free, non-empty alternating sequence
of nodes and edges such that each edge is connected to each subsequent node,
and the end node is the same as the start node \cite{diestel}.

\section{Motivation}
According to Watts and Strogatz \cite{watts}, small-world graphs are the norm
for real-world data.
A small-world graph is defined as a tightly-clustered graph with small path
lengths \cite{watts}. It is easy to see that a tightly-clustered graph with
small path lengths will contain many triangles. The number of triagles are thus
important for understanding the data in a graph because they give us an idea of
the cliquishness of a graph.

The standard metrics for the cliquishness of a graph are the clustering
coefficient and the transitivity ratio, which have a high correlation to the
number of triangles in a graph. The clustering coefficient of a
node is the fraction of possible edges it could have to other nodes divided by
the number of edges it actually has \cite{watts}. One of the ways of
calculating the global clustering coefficient is actually by calculating the
number of triangles, multiplying that number by three, and dividing by the
number of connected triples of vertices \cite{newman}.
The transitivity of a graph (which is the transitivity ratio restricted to
undirected graphs) is another term for the global clustering
coefficient \cite{schank:cluster}.

\section{Triangle Counting and Listing Works}

\subsection{Exact counting and listing algorithms}
Schank and Wagner present an overview of triangle algorithms. They divide
triangle algorithms into either counting or listing algorithms, where counting
algorithms simply produce the number of triangles, while listing algorithms
produce a list of every triangle present in the graph \cite{schank:counting}.

They present several exact counting methods in their paper. The simplest
traverses
the adjacency matrix using the matrix multiplication solution to the shortest
paths problem to get a count of the triangles, with a runtime of $O(n^3)$
\cite{schank:counting}. A slightly better exact counting algorithm, which they
call \textit{node-iterator}, traverses all nodes and counts every
set of three connected nodes and obtains a runtime of $O(nd_{max}^2)$.

The fastest currently-known exact triangle counting algorithm is known as AYZ
for the authors Alon, Yuster, and Zwick, which uses fast matrix multiplication
to traverse the nodes and count all triangles and has a runtime of $O(m^{3/2})$
\cite{ayz}. Thus, the fastest exact triangle counting takes exponential time,
which is why \cite{original} presents a fast estimation algorithm.

\subsection{Streaming algorithms}
A streaming algorithm is an algorithm that handles data that arrives in a
data stream. Bar-Yossef et. al. have developed a streaming algorithm to
approximate the number of triangles in a large graph. This algorithm has the
advantage of being space efficient and only requiring a single pass over the
data \cite{baryossef}. 

\subsection{Semi-streaming algorithms}
Becchetti, Boldi, and Castillo present an algorithm to estimate the local
number of triangles in a graph (for each node, compute the number of triangles
that the node participates in). They chose to use a semi-streaming algorithm
for this because a streaming algorithm would constrain the memory usage too
much to be useful. The semi-streaming algorithm limits the passes over the data
to be at most $O(\log N)$ \cite{becchetti}.

\subsection{Algorithms that minimize disk I/O}
Triangle listing algorithms suffer from huge datasets and the fact that
in-memory algorithms are unable to handle looking at each node without
significant slowdowns from disk I/O. Chu and Cheng have solved this problem
using an algorithm that is designed to work on neighboring vertices as a whole
in order to minimize disk I/O accesses \cite{chu}.

\subsection{Algorithms that use MapReduce and Hadoop}
Exact triangle counting also has the issue of huge graphs and many disk I/O
accesses. Suri and Vassilvitskii have developed an algorithm to calculate the
number of triangles using MapReduce on a computing cluster in order to mitigate
this issue \cite{suri}.

The ability to massively parallelize triangle counting is important due to the
size of the graphs being mined, and the algorithm presented in \cite{original} 
is able to be run on a MapReduce system as well.

\section{Background Works}
This section will introduce the topics needed to understand the theorems and
algorithms \cite{original} presents.

\subsection{Adjacency matrices}
An adjacency matrix of a graph which has n nodes is a matrix $A = 
(a_{ij})_{nxn}$ which is defined by $a_{ij} = 1$ if $v_iv_j \in E$ and $0$
otherwise
\cite{diestel}.  This gives us a way to do linear algebra operations on graphs.

\subsection{Eigenvalues}
For a matrix A, given the expression $Ax = \lambda x$, where $x$ is a vector and
$\lambda$ is a real or complex number, $x$ is said to be the eigenvector of $A$
and $\lambda$ is said to be its eigenvalue. \cite{lovasz}

\subsection{Estimating eigenvalues}
The Lanczos method is an iterative method of estimating the eigenvalues in a
matrix. It provides a good, fast estimate for large graphs with sparse
eigenvalues \cite{golub}.

\subsection{Power laws}
A power law is a J-shaped, highly skewed distribution function of some
empirical data with a long tail \cite{simon}.

A degree power law is a graph with a power law for graph degrees. In other
words, it has a few nodes with a high degree and many nodes with a low degree
\cite{mihail}.

If a graph has a degree power law, it also has an eigenvalue power law
\cite{mihail}. Degree power laws are common in real networks \cite{original},
and so by extension eigenvalue power laws are also common. The fact that
these graphs have eigenvalue power laws makes the eigenvalues fast to calculate
because there are only a few important eigenvalues, and many unimportant
(small) eigenvalues, so the Lanczos method can be used to find the most
important eigenvalues and thus find a close approximation quickly \cite{original}.

\subsection{Number of closed walks in a graph}
In \cite{harary}, Harary and Schwenk prove that the number of closed walks of
length n in a graph is the sum of the nth powers of the graph's eigenvalues. By
extension, the number of closed walks of length 3 is the sum of cubes of the
graph's eigenvalues, and a closed walk of length 3 is a triangle.

\section{Method}
The main idea of \cite{original} is the fact that the total number of triangles
in a graph is proportional to the sum of cubes of its adjacency matrix
eigenvalues. The formula developed by \cite{original} is the following:
$$\Delta(G) = \frac{1}{6} \sum_{i=1}^n \lambda_i^3$$

The proposed algorithm EigenTriangle which calculates the number of triangles
involves
iteratively using the Lanczos method to calculate the eigenvalues, cubing and
summing them as we calculate them.

In addition, \cite{original} presents another formula and algorithm which are
extensions of EigenTriangle that count the number of triangles $\Delta_i$ that
node i participates in, calling these EigenTriangleLocal. The formula for this
is the following: $$\Delta_i = \frac{\sum_j \lambda_j^3 u_{i,j}^2}{2}$$
Similarly, the
algorithm for this works by using the Lanczos method to compute the cubes of the
eigenvalues of the adjacency matrix.

This method provides a mean speedup of 250x compared to the aforementioned
\textit{node-iterator} algorithm presented by \cite{schank:counting}.

In addition to the algorithms developed, \cite{original} also uncovers a
power law between a graph's average degree and the number of triangles present
in the graph. \cite{original} calls this the DEGREE-TRIANGLE power law. 

\cite{original} also notes that the idea of using eigenvalues to calculate
the number of triangles in a
graph also works for Kronecker graphs and Erdos-Renyi graphs.

\section{Conclusion}
This has been an explanation and related works for \cite{original}. This paper
has explained the basic concepts, motivation, background and related works so
that \cite{original} is easier to understand.

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
\end{document}
